### 2017-05-31 doc/categories.tex
discussion of results in section 4.3 expanded.

Models that utilize the sequential structure of language 
have the capacity to interpret the same word-type differently depending on
the context. The omission score distributions in Section \ref{sec:omitimaginet} 
show that in the case of {\sc Imaginet} the 
pathways are differentially sensitive to content vs.\ function
words. In principle, this may be either just due to purely lexical features or the model 
may actually learn to pay more attention to the same word type in appropriate
contexts. This section investigates to what extent {\sc Visual} and {\sc Textual}
discriminate between occurrences of a given word in different positions and 
grammatical functions. The analysis we described here takes \label{edit:beyonlexicalgeneral}
the omission scores as input data, therefore it can be potentially applied 
to any architecture for which the omission scores can be computed. However,
the presented analysis and results regarding word positions can only be meaningful
for Recurrent Neural Networks as they compute their representations sequentially and are not
limited by fixed window sizes.\footnote{CNNs with multi-word filters
and tree-structured recursive neural networks do not incrementally build representations
of sentences in a left-to-right or right-to-left fashion. 
Bi-directional RNNs, on the other hand, are sensitive by word-order and can potentially
learn to handle the same word in different positions differently. \label{edit:foot}}

We fit four L2-penalized linear regression models which predict the omission 
scores per token with the following predictor variables: 
\begin{itemize}
	\item {\sc LR word}: word type
	\item {\sc LR +dep}: word type, dependency label and their interaction 
	\item {\sc LR +pos}: word type, position (binned as {\sc first, second, third, middle,
	antepenult, penult, last}) and their interaction
	\item {\sc LR full}: word type, dependency label, position, word:dependency interaction, 
	word:position interaction
\end{itemize}

\noindent We split the 5000-image portion of MSCOCO validation data into two equal parts, fit the 
models on the first part, and compute the 
proportion of variance explained on the second part. 
For comparison we also use the {\sc Sum} model which composes word
embeddings via summation, and uses the same loss function as {\sc
  Visual}. This model is unable to encode information about word
order, and thus is a good baseline here as we investigate the
sensitivity of the networks to positional and structural cues.

@@

\todo[inline]{I think we can improve the discuss of results below.}
Table~\ref{tab:lr-r2} shows the proportion of variance $R^2$ in omission
scores explained by the linear regression with the different predictors.
Figure~\ref{fig:rsquared} offers a different view of the data, showing
the increase or decrease in $R^2$ for the models relative to {\sc
  LR~+pos}.  
The results reveal that adding additional information on
top of lexical features in the case of {\sc Sum} does increase the
explained variance slightly, which is most likely due to the unseen words
in the held out set. More interestingly, there is a sizeable
increase in $R^2$ between {\sc LR~+pos} and {\sc LR~full} in
the case of {\sc Visual}, suggesting that 
the omission scores in the case of {\sc Visual} depend on the words'
grammatical function in sentences, {\it even after controlling for word
identity and linear position.}

The raw $R^2$ scores show that for the {\sc Textual} and {\sc LM}
models the word-type predicts the omission-score to much smaller
degree compared to {\sc Visual}, and the increase in $R^2$ shows that
the position of the word adds considerable amount of information for
these language models: this is not surprising considering that the omission
scores are measured with respect to the final activation state, and
given the fact that in a language model the recent history is most
important for accurate prediction. Less expectedly, for the {\sc Visual} model,
dependency labels are more informative than linear position,
hinting at the importance of syntactic structure beyond linear order.

Overall, when regressing on word identities, word position and
dependency labels, the {\sc Visual} model's omission scores are the
hardest to predict of the four models suggesting that the model may be
encoding additional structural features not captured by these
predictors.

>>>>

Models that utilize the sequential structure of language 
have the capacity to interpret the same word-type differently depending on
the context. The omission score distributions in Section \ref{sec:omitimaginet} 
show that in the case of {\sc Imaginet} the 
pathways are differentially sensitive to content vs.\ function
words. In principle, this may be either just due to purely lexical features or the model 
may actually learn to pay more attention to the same word type in appropriate
contexts. This section investigates to what extent our models
discriminate between occurrences of a given word in different positions and 
grammatical functions. The analysis we described here takes \label{edit:beyonlexicalgeneral}
the omission scores as input data, therefore it can be potentially applied 
to any architecture for which the omission scores can be computed. However,
the presented analysis and results regarding word positions can only be meaningful
for Recurrent Neural Networks as they compute their representations sequentially and are not
limited by fixed window sizes.\footnote{CNNs with multi-word filters
and tree-structured recursive neural networks do not incrementally build representations
of sentences in a left-to-right or right-to-left fashion. 
Bi-directional RNNs, on the other hand, are sensitive by word-order and can potentially
learn to handle the same word in different positions differently. \label{edit:foot}}

We fit four L2-penalized linear regression models which predict the omission 
scores per token with the following predictor variables: 
\begin{enumerate}
	\item {\sc LR word}: word type
	\item {\sc LR +dep}: word type, dependency label and their interaction 
	\item {\sc LR +pos}: word type, position (binned as {\sc first, second, third, middle,
	antepenult, penult, last}) and their interaction
	\item {\sc LR full}: word type, dependency label, position, word:dependency interaction, 
	word:position interaction
\end{enumerate}

\noindent We use the 5000-image portion of MSCOCO validation data for 
training and test. The captions contain about 260,000 words in total, of 
which we use 100,000 to fit the regression models. We then use the rest 
of the words to compute the proportion of variance explained by the models. 
For comparison we also use the {\sc Sum} model which composes word
embeddings via summation, and uses the same loss function as {\sc
  Visual}. This model is unable to encode information about word
order, and thus is a good baseline here as we investigate the
sensitivity of the networks to positional and structural cues.

@@

Table~\ref{tab:lr-r2} shows the proportion of variance $R^2$ in omission
scores explained by the linear regression with the different predictors.
The raw $R^2$ scores show that for the language models {\sc LM} and 
{\sc Textual}, the word-type predicts the omission-score to much smaller 
degree compared to {\sc Visual}. Moreover, adding information about 
either the position or the dependency labels increases the explained variance for all models. 
However, for the {\sc Textual} and {\sc LM} models the position of the word adds 
considerable amount of information. This is not surprising considering that the omission
scores are measured with respect to the final activation state, and
given the fact that in a language model the recent history is most
important for accurate prediction.

Figure~\ref{fig:rsquared} offers a different view of the data to, showing
the increase or decrease in $R^2$ for the models relative to {\sc LR~+pos} 
to emphasise the importance of syntactic structure beyond the position in the sentence.
Interestingly, for the {\sc Visual} model, dependency labels are 
more informative than linear position, hinting at the importance of syntactic 
structure beyond linear order. There is a sizeable increase in $R^2$ between 
{\sc LR~+pos} and {\sc LR~full} in the case of {\sc Visual}, suggesting that 
the omission scores for {\sc Visual} depend on the words'
grammatical function in sentences, {\it even after controlling for word
identity and linear position.}  In contrast, adding additional information on
top of lexical features in the case of {\sc Sum} increases the
explained variance only slightly, which is most likely due to the unseen words
in the held out set.

Overall, when regressing on word identities, word position and
dependency labels, the {\sc Visual} model's omission scores are the
hardest to predict of the four models suggesting that the model may be
encoding additional structural features not captured by these predictors. 
We will look more deeply into such potential features in the following sections.
